The Impact of Alternation

### 2017-05-07 draft.tex
added predicate abstraction algo

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq 0$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading $a \in \Sigma$.

We generalize the post-image function to finite sequences of input
events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
\rightarrow \bot)$ is satisfiable ? A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains of formulae\footnote{For instance,
  when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq 0,
  q \wedge x \teq 1, \ldots$}.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it simply an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota$, in which the data variables are unconstrained
and over-approximates the set of reachable valuations (\ref{it1:inv}),
while excluding those valuations that satisfying the acceptance
condition (\ref{it2:inv}).

@@

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
defines\footnote{ A formula $\phi(Q,\vec{x})$ defines $\Inv$ iff for
  all $\beta: Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow
  \Data^\I$: $\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. } a
safety invariant for $\A$. Although this condition can be effectively
checked using a decision procedure for the theory $\theory(\sig,\I)$,
there is no guarantee that this check will ever succeed.

The solution we adopt is the use of \emph{abstraction} to ensure
termination of invariant computations. However, it is worth pointing
out from the start that abstraction will allow us to build invariants
which are not necessarily safety invariants. To converge towards
meeting the latter condition, we shall resort to \emph{counterexample
  guided abstraction refinement} (CEGAR).

@@

In general, given an unsatisfiable conjunction $\pi \equiv
\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
  \in X}$ of any sort, a solution of the \emph{interpolation problem}
$\pi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
\item $I_0 \equiv \top$,  
%
\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
  I_i[X_i/X]$, for all $i \in [1,n]$, and
%
\item $I_n \equiv \bot$. 
\end{inparaenum}
A classical method for abstraction refinement is to add the elements
of the interpolant obtained from a proof of spuriousness to the set of
predicates. This guarantees progress, meaning that the particular
spurious counterexample, from which the interpolant was generated,
will never be revisited in the future. In many practical test cases,
this progress property eventually yields a safety invariant.

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: \[\pi(u)
\equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup
\vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)\] where
$Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped sets of
boolean variables corresponding to the set $Q$ of states of $\A$. The
first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the initial
configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$ replaced
by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
\emph{replacement sets} $R_\ell \in Q_\ell$, $\ell\in [0,n]$, which
are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\iota(Q_0)}$, 
%
\item for each $\ell\in[1,n]$: $\theta_\ell \equiv
  \bigwedge_{q_{\ell-1}\in R_{\ell-1}} (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$.
\end{compactitem}
Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
\theta_n$ simulates the run of $\A$ over $u$ and $\theta_{n+1}$ is the
part of the acceptance condition which forces the last remaining
non-final states to be false.

We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
\phi_1(Q,\vec{x}_0\cup\vec{x}_1), \ldots,
\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$, where $\phi_0$ is the
initial configuration $\iota$ and for each $k>0$, $\phi_k$ is obtained
from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
transition function of $\A$. The main reason behind the above
definition of the interpolation problem is that we would like to
obtain an interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with
  the data constraints that must hold locally}, whenever the control
reaches a certain boolean configuration. This association of states
with data valuations is tantamount to defining efficient
semi-algorithms, based on lazy abstraction \cite{HJMS02}. Observe
that, because the states are replaced with transition formulae when
moving one step in a run, the formulae thus produced lose track of the
control history and are not directly suitable to producing
interpolants that can be applied locally. Moreover, the abstraction
defined by the interpolants generated in this way can
\emph{over-approximate the control structure} of an automaton, in
addition to the sets of data values encountered throughout its runs.

The correctness of this interpolation-based abstraction refinement
setup is captured by the progress property below, which guarantees
that adding the formulae of an interpolant for $\pi(u)$ to the set
$\preds$ of predicates suffices to exclude $u$ from future searches.

\begin{lemma}\label{lemma:progress}
  Given a non-empty sequence of input events $u \in \Sigma^*$ such
  that $\Accept{\A}(u)$ is unsatisfiable, then the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\pi(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_n,\bot}$ is an
    interpolant for $\pi(u)$ and $I_i \in \preds$, for all
    $i\in[1,n]$, then $\AbsAccept{\A}(u)$ is unsatisfiable.
  \end{compactenum}
\end{lemma}
\proof{

\qed}

>>>>

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

@@

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
effect of reading the event $a$.

We generalize the post-image function to finite sequences of input
events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
\rightarrow \bot)$ is satisfiable ?  Observe that, since we ask a
satisfiability query, the final states of $\A$ need not be
constrained\footnote{ Since each state occurs positively in
  $\Accept{\A}(u)$, this formula has a model iff it has a model with
  every $q \in F$ set to true.}. A na\"ive semi-algorithm enumerates
all finite sequences and checks the satisfiability of $\Accept{\A}(u)$
for each $u \in \Sigma^*$, using a decision procedure for the theory
$\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain-based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains, e.g.\ $q \wedge x \teq 0, q \wedge x
\teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
$\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. If $\Inv$
satisfies only the first point above, we call it simply an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, in which the
data variables are unconstrained, and over-approximates the set of
reachable valuations (\ref{it1:inv}), while excluding those valuations
which satisfy the acceptance condition (\ref{it2:inv}).

@@

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
defines a safety invariant for $\A$. Although this condition can be
effectively checked using a decision procedure for the theory
$\theory(\sig,\I)$, there is no guarantee that this check will ever
succeed.

The solution we adopt is the use of \emph{abstraction} to ensure
termination of invariant computations. However, it is worth pointing
out from the start that abstraction will allow us to build invariants
which are not necessarily safety invariants. To meet the latter
condition, we resort to \emph{counterexample guided abstraction
  refinement} (CEGAR).

@@

In general, given an unsatisfiable conjunction $\pi \equiv
\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
  \in X}$ of any sort, a solution of the \emph{interpolation problem}
$\pi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
\item $I_0 \equiv \top$,  
%
\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
  I_i[X_i/X]$, for all $i \in [1,n]$, and
%
\item $I_n \equiv \bot$. 
\end{inparaenum}
A classical method for abstraction refinement is to add the elements
of the interpolant obtained from a proof of spuriousness to the set of
predicates. This guarantees progress, meaning that the particular
spurious counterexample, from which the interpolant was generated,
will never be revisited in the future. Though not always, in many
practical test cases this progress property eventually yields a safety
invariant.

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{equation}\label{eq:interpolation-problem}
\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)
\end{equation} 
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$, 
%
\item for each $\ell\in[1,n]$: $\theta_\ell \equiv
  \bigwedge_{q_{\ell-1}\in R_{\ell-1}} (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$.
\end{compactitem}
Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
\theta_n$ simulate the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.

We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
\phi_1(Q,\vec{x}_0\cup\vec{x}_1), \ldots,
\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$, where $\phi_0$ is the
initial configuration $\iota$ and for each $k>0$, $\phi_k$ is obtained
from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
transition function of $\A$. The main idea behind the above definition
of the interpolation problem is that we would like to obtain an
interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with the
  data constraints that must hold locally}, whenever the control
reaches a certain boolean configuration. This association of states
with data valuations is tantamount to defining efficient
semi-algorithms, based on lazy abstraction \cite{HJMS02}. Observe
that, because the states are replaced with transition formulae when
moving one step in a run, the formulae thus produced lose track of the
control history and are not directly suitable to producing
interpolants that can be applied locally. Furthermore, the abstraction
defined by the interpolants generated in this way can also
\emph{over-approximate the control structure} of an automaton, in
addition to the sets of data values encountered throughout its runs.

The correctness of this interpolation-based abstraction refinement
setup is captured by the progress property below, which guarantees
that adding the formulae of an interpolant for $\pi(u)$ to the set
$\preds$ of predicates suffices to exclude the spurious counterexample
$u$ from future searches.

\begin{lemma}\label{lemma:progress}
  Given a non-empty sequence of input events $u \in \Sigma^*$ such
  that $\Accept{\A}(u)$ is unsatisfiable, then the following
  hold: \begin{compactenum}
  \item\label{it1:progress} $\pi(u)$ is unsatisfiable, and
    %
  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_{n+1},\bot}$
    is an interpolant for $\pi(u)$ and $\set{I_i \mid i \in[1,n+1]}
    \subseteq \preds$ then $\AbsAccept{\A}(u)$ is unsatisfiable.
  \end{compactenum}
\end{lemma}
\proof{ Let $u = a_1\ldots a_n$ and  $\pi(u) \equiv \theta_0(Q_0) \wedge
  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
  \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge
  \theta_{n+1}(Q_n)$ in the following.

  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
  data valuations for $Q_0,\ldots,Q_n$ and
  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
  $\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
  all $q \in Q$. For the induction step $n > 0$, we assume that
  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
  $\pi(u)$, we have: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]

  But since $\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
  \theta_n(Q_{n-1}\cup Q_n, \vec{x}_{n-1} \cup \vec{x}_n)$, we have
  that $\beta_n,\vec{x}_{n-1} \cup \vec{x}_n \models
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  for every $q \in Q$ such that $\beta_{n-1}(q_{n-1}) = \true$. Then
  we have $\beta,\vec{x}_0\cup\ldots\cup\vec{x}_n \models \phi_n$
  where $\beta(q) = \beta_n(q_n)$, for all $q \in Q$ and $\phi_n$ is
  obtained from $\phi_{n-1}$ by replacing each boolean variable $q$ by
  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$. Observe
  that, because each boolean variable occurs positively in
  $\phi_{n-1}$, the truth value of
  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$
  under valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
  $q \in Q$ such that $\beta(q) = \false$. We have thus obtained a
  model for $\phi_n$ and thus for $\Post{\A}(u)$. Since moreover,
  $\theta_{n+1}[Q/Q_n] = \true$ in this model, we obtain a model for
  $\Accept{\A}(u)$. Thus $\pi(u)$ is satisfiable implies that
  $\Accept{\A}(u)$ is satisfiable, which proves the statement of point
  (\ref{it1:progress}).

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, for all $k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{rcll}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \text{ def. of $\AbsPost{\A}$} \\
    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \text{ ($\Asterisk$)} \\
    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \text{ by ind. hyp.} \\ 
    & \models & I_n[Q_n/Q] & \hfill\text{\qed}
    \end{array}\]
    We prove the final point ($\Asterisk$). If
    $\beta_n\cup\nu_{n-1}\cup\nu_n \models
    \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q]$,
    for some valuations $\beta_{n-1} : Q_{n-1} \rightarrow \booli$ and
    $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
    build a valuation $\beta_{n-1}$ such that $\beta_{n-1}\cup\beta_n
    \cup \nu_{n-1}\cup\nu_n \models \AbsPost{\A}(\iota,a_1\ldots
    a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$, as follows. For each
    occurrence of a subformula $\Delta(q,a_n)$ in
    $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
    $\beta_{n-1}(q_{n-1}) = \true$ if $\beta_n\cup\nu_{n-1}\cup\nu_n
    \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
    otherwise. Since there are no negated occurrences of such
    subformulae, the definition of $\beta_{n-1}$ is consistent, and
    the check $\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
    \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
    is straightforward. \qed}

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Although termination is not guaranteed,
  we call it an algorithm for conciseness.}  \ref{alg:predabs} builds
an \emph{abstract reachability tree} (ART) whose edges are labeled
with input events and each node $n$ is labeled with the formula
$\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
labeling the path from the root of the tree to $n$. It uses a covering
relation $\lhd$ between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample $u$ is eliminated by computing the
interpolants corresponding to the least unfeasible suffix $v$ of $u$
and recomputing the subtree of the ART rooted in the initial node of
$v$.

\begin{algorithm}[t!]
\begin{algorithmic}[0]
\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
over the alphabet $\Sigma$ of input events

\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
\in L(\A)$ otherwise
\end{algorithmic}

\begin{algorithmic}[1]
  \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, where $N \leftarrow
  \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
  \set{(\rootNode,\iota)}$ and $\lhd \leftarrow \emptyset$

  \While {$\worklist \neq \emptyset$}

  \State dequeue $n$ from $\worklist$

  \State $N \leftarrow N \cup \set{n}$

  \If{$\Lambda(n) \wedge \bigwedge_{q \in Q \setminus F} (q
    \rightarrow \bot)$ is satisfiable} \Comment{counterexample found}

  \State let $u = a_1\ldots a_k$ be the label of the path from
  $\rootNode$ to $n$ in $\Art$

  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{counterexample is
    real}

  \State let $\beta \cup \nu_1 \cup \ldots \cup \nu_k \models
  \Accept{\A}(u)$, where $\beta : Q \rightarrow \booli$,
  $\nu_1,\ldots,\nu_k : \vec{x} \rightarrow \Data^\I$

  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
  \Comment{$w \in L(\A)$ by construction}

  \Else \Comment{spurious counterexample}

  \State let $p \leftarrow \Call{FindPivot}{u,\Art}$

  \State let $v$ be the label of the path from $p$ to $n$ in $\Art$

  \State $\preds \leftarrow \preds \cup \set{I_1,\ldots,I_\ell}$,
  where $\tuple{\top,I_1,\ldots,I_\ell,\bot}$ is the interpolant for
  $\overline{\pi}(v)$

  \State let $\SubArt = \tuple{N',E',\Lambda',\lhd'}$ be the subtree
  rooted at $p$ in $\Art$

  \For{$(m,q) \in \lhd$ such that $q \in N'$}

  \State remove $m$ from $N$ and enqueue $m$ into $\worklist$

  \State remove $\SubArt$ from $\Art$

  \State enqueue $p$ into $\worklist$

  \EndFor

  \EndIf

  \Else

  \For{$a \in \Sigma$}

  \State let $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$

  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$}

  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$n$ covers $m$}

  \Else

  \State let $s \leftarrow \Call{NewNode}{}$ and $\Lambda \leftarrow
  \Lambda \cup \set{(s,\phi)}$

  \State $R \leftarrow \set{m \in \worklist \mid \Lambda(m) \models
    \phi}$ \Comment{worklist nodes covered by $s$}

  \For{$r \in R$}

  \For{$m \in N$ such that $(m,r) \in E$}

  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered children from $R$ into $s$}

  \EndFor

  \For{$(m,r) \in \lhd$}

  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered nodes from $R$ into $s$}

  \EndFor

  \EndFor

  \State remove $R$ from $\Art$

  \State enqueue $s$ into $\worklist$ \Comment{recompute the subtree rooted at $s$}

  \EndIf

  \EndFor

  \EndIf

  \EndWhile

  \State {\bf return} $\true$
\end{algorithmic}
\caption{Lazy Predicate Abstraction for Checking Emptiness of ADA}
\label{alg:predabs}
\end{algorithm}
